---
title: "ML Project"
author: "CR, CT, FF"
format: html
editor: visual
---

# ML PROJECT

```{r}

set.seed(1)
O_DS <- readRDS("/Users/francescofavagrossa/Desktop/Machine Learning/ML Project/Dataset/TicTacToe_dataset.RDS")
library(caTools); library(caret); library(corrplot); library(DiceKriging); library(dplyr)
library(e1071); library(factoextra); library(farff); library(GauPro); library(ggplot2)
library(gridExtra); library(mlr); library(mlrMBO); library(parallelMap); library(tibble)
library(tidyverse); library(glmnet); library(tidyr)
```

## 1. Analisi esplorativa e preparazione dei dati

```{r}

colSums(is.na(O_DS))
ttcorr <- cor(O_DS)
corrplot(ttcorr, method = "circle", type = "upper", 
         order = "hclust", tl.col = "black", tl.srt = 45, 
         title = "Features Correlations Matrix ", mar=c(0,0,1,0))
matrice_correlazione <- cor(O_DS)

dati_corr_long <- matrice_correlazione %>%
  as.data.frame() %>%
  rownames_to_column(var = "Var1") %>%
  pivot_longer(
    cols = -Var1,
    names_to = "Var2",
    values_to = "Correlation"
  )
dati_corr_long$Var1 <- factor(dati_corr_long$Var1, levels = unique(dati_corr_long$Var1))
dati_corr_long$Var2 <- factor(dati_corr_long$Var2, levels = unique(dati_corr_long$Var2))
grafico_heatmap <- ggplot(dati_corr_long, 
                          aes(x = Var1, y = Var2, fill = Correlation)) +
  
  geom_tile() + 
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 3) +
  scale_fill_gradient2(
    low = "skyblue", 
    high = "skyblue3", 
    mid = "white",
    midpoint = 0,
    limit = c(-1, 1),
    name = "Correlation"
  ) +
  labs(
    title = "Correlation heatmap for Features and Class",
    x = "Variable",
    y = "Variable"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

print(grafico_heatmap)

tt_noclass <- O_DS[,1:12]
pca_result <- prcomp(tt_noclass, center = TRUE, scale. = TRUE) # PCA
p2 <- fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50)) +
  theme(plot.title = element_text(face = "bold"))

grid.arrange(arrangeGrob(grafico_heatmap, bottom = "(a) Opt. Path "),
             arrangeGrob(p2, bottom = "(b) Confusion Matrix"), ncol = 2, nrow = 1)

```

## **2. K-Nearest Neighbors**

Trasformiamo la classe in factor e creiamo la divisione in dataset e testset, tenendo conto della distribuzione della variabile target.

```{r}
ds <- O_DS
ds$class <- as.factor(ds$class)
ixs <- createDataPartition(ds$class, time = 1, p = 0.9) 
dataset <- ds[ixs$Resample1, ] # 90%
testset <- ds[-ixs$Resample1, ] # 10%. from now on we treat this test set as if it was the set we do not have.

featRengers <- apply(dataset[, -ncol(dataset)], 2, range) 

for( j in 1:ncol(featRengers)){
  dataset[,j] <- (dataset[,j] - featRengers[1,j])/diff(featRengers[,j])
}

for( j in 1:ncol(featRengers)){
  testset[,j] <- (testset[,j] - featRengers[1,j])/diff(featRengers[,j])
}
```

### 2.1 Hold-out with

with grid search.

```{r}

k <- seq(from = 1, to = 51, by = 2)
knn_k <- matrix(ncol = 2, nrow = length(k))
knn_k[,1] <- t(k)

trnIxs <- createDataPartition(dataset$class, time = 1, p = 0.7)

# Definisco i subset per chiarezza
train_sub <- dataset[trnIxs$Resample1, -ncol(dataset)]
valid_sub <- dataset[-trnIxs$Resample1, -ncol(dataset)]
cl_train_sub <- dataset$class[trnIxs$Resample1]
cl_valid_sub <- dataset$class[-trnIxs$Resample1]

for(i in 1:length(k)){ 
  knn <- knn_k[i,1]
  preds <- knn3Train(train = train_sub, 
                   test = valid_sub,
                   cl = cl_train_sub, 
                   k = knn, 
                   prob = F, 
                   use.all = T)
  
  knn_k[i,2] <- 1-(mean(cl_valid_sub != preds)) 
}

knn_df <- data.frame(k_values = knn_k[, 1], accuracy = knn_k[, 2])
ggplot(knn_df, aes(x = k_values, y = accuracy)) +
    geom_line(color = "blue") +
  geom_point(color = "black") +
  labs(
    title = "Accuracy for every odd value of k from 1 to 51 using HOLD OUT",
    x = "k values",
    y = "accuracy"
  ) +
  theme_minimal()

index_opt <- as.numeric(which.max(knn_k[,2]))
k_opt <- knn_k[index_opt,1]

preds_test <- knn3Train(train = dataset[, -ncol(dataset)], 
                        test = testset[, -ncol(testset)],
                        cl = dataset$class, 
                        k = k_opt, 
                        prob = FALSE, 
                        use.all = TRUE)
preds_test <- factor(preds_test, levels = levels(testset$class))

confMat <- confusionMatrix(preds_test, testset$class)
conf_mat_df <- as.data.frame(confMat$table)
colnames(conf_mat_df) <- c("Predicted", "Actual", "Freq")

class_levels <- levels(testset$class) 
conf_mat_df$Actual    <- factor(conf_mat_df$Actual,    levels = rev(class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = class_levels)

ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") +
  coord_fixed() +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) +
  labs(
    title = paste0("Confusion matrix KNN - HOLD OUT (k = ", k_opt, ") - Test Set"),
    x = "Predicted class",
    y = "Actual class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right")
```

### 2.2 K-fold cross validation

```{r}

control <- trainControl(method = "repeatedcv",
                        number = 10,          
                        repeats = 3,           
                        classProbs = TRUE,     
                        summaryFunction = defaultSummary) 
k_values <-  seq(from = 1, to = 51, by = 2) 
levels(dataset$class) <- make.names(levels(dataset$class))
knn_model <- caret::train(class ~ .,         
                          data = dataset,       
                          method = "knn",             
                          tuneGrid = expand.grid(k = k_values), 
                          trControl = control,        
                          metric = "Accuracy")  
miglior_k <- knn_model$bestTune  
cat("\nIl valore di k ottimale (con la migliore Accuracy) è:", miglior_k$k, "\n")
knn_results_df <- data.frame(
  k_values = knn_model$results$k,
  accuracy = knn_model$results$Accuracy
)
ggplot(knn_results_df, aes(x = k_values, y = accuracy)) +
  geom_line(color = "blue") +
  geom_point(color = "black") +
  labs(
    title = "Accuracy for every odd values of k from 1 to 51 using 10FCV.",
    x = "k values",
    y = "accuracy"
  ) +
  theme_minimal()

predictions <- predict(knn_model, newdata = testset)
levels(testset$class) <- make.names(levels(testset$class))
conf_matrix <- confusionMatrix(predictions, testset$class)

conf_mat_df <- as.data.frame(conf_matrix$table)
colnames(conf_mat_df) <- c("Predicted", "Actual", "Freq")

class_levels <- levels(testset$class) 

conf_mat_df$Actual <- factor(conf_mat_df$Actual, levels = rev(class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = class_levels)

ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Frequenza") +
  coord_fixed() +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) + 
  labs(
    title = paste0("Confusion matrix for KNN - 10FCV (k = ", miglior_k$k, ") - Test Set"),
    x = "Predicted class",
    y = "Actual class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

```

### 2.3 Bo

```{r}

par.set <- makeParamSet(makeIntegerParam("k", lower = 1, upper = 51))

task <- makeClassifTask(data = dataset , target = "class")
ctrl <- makeMBOControl()
ctrl <- setMBOControlTermination(ctrl, iters = 30)
tune.ctrl <- makeTuneControlMBO(mbo.control = ctrl)

lrn <- makeLearner("classif.knn")

# tuning
run <- tuneParams(
  learner     = lrn,
  task        = task,
  resampling  = cv3,
  measures    = acc,
  par.set     = par.set,
  control     = tune.ctrl,
  show.info   = TRUE
)

ys <- getOptPathY(run$opt.path)
df <- data.frame(
  iter = seq_along(ys),
  value = ys,
  best_so_far = cummax(ys)
)
ggplot(df, aes(x = iter)) +
  geom_line(aes(y = value), color = "gray", linewidth = 1.2) +
  geom_point(aes(y = value), color = "black", size = 2) +
  geom_step(aes(y = best_so_far), color = "blue", linewidth = 1.4) +
  labs(
    title = "Optimization Path KNN - BO",
    x = "Iteration",
    y = "Objective Value"
  ) +
  theme_minimal(base_size = 14)

ks_df <- as.data.frame(getOptPathX(run$opt.path))
iters <- seq_len(nrow(ks_df))
accs <- getOptPathY(run$opt.path)

dfk <- data.frame(
  accuracy = ys,
  k = ks_df$k
)
dfk_sorted <- dfk %>% arrange(k)

ggplot(dfk_sorted, aes(x = k)) +
  geom_line(aes(y = accuracy), color = "blue", linewidth = 1.2) +
  geom_point(aes(y = accuracy), color = "black", size = 2) +
  labs(
    title = "Accuracy for every odd value of k from 1 to 51 using 10FCV. ",
    x = "Value of K",
    y = "Accuracy"
  ) +
  theme_minimal(base_size = 14)

final_learner <- setHyperPars(lrn, par.vals = run$x)
final_model <- mlr::train(final_learner, task)

preds <- predict(final_model, 
                 newdata = testset)$data$response

levels(testset$class) <- make.names(levels(testset$class))
actual_classes <- testset$class

conf_matrix <- confusionMatrix(preds, actual_classes)

conf_mat_df <- as.data.frame(conf_matrix$table)
colnames(conf_mat_df) <- c("Predicted", "Actual", "Freq")

current_class_levels <- levels(actual_classes)
conf_mat_df$Actual <- factor(conf_mat_df$Actual, levels = rev(current_class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = current_class_levels)
miglior_k_bo <- run$x$k

ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  # Scala di colore
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") +
  coord_fixed() +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) +
  labs(
    title = paste0("Confusion matrix KNN - BO (k = ", miglior_k_bo, ") - Test Set"),
    x = "Predicted class",
    y = "Actual class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"), # Titolo in grassetto
    legend.position = "right"
  )

```

## 3. Support Vectors Machine

### 3.1 K-fold cross validation

```{r}

dataset <- ds[ixs$Resample1, ]
testset <- ds[-ixs$Resample1, ]
c_values <- 10^(-1:2)   
gamma_values <- 10^(-2:1)

tuned_mdl <- tune.svm(
  class ~ .,              
  data = dataset,        
  kernel = "radial",     
  cost = c_values,        
  gamma = gamma_values,   
  tunecontrol = tune.control(cross = 10) 
)

best_C <- tuned_mdl$best.parameters$cost
best_Gamma <- tuned_mdl$best.parameters$gamma

mdl_final <- svm(
  class ~ .,              
  data = dataset,        
  type = "C-classification",
  kernel = "radial",      
  cost = best_C,       
  gamma = best_Gamma,  
  scale = FALSE           
)

preds_test <- predict(mdl_final, testset)

levels(testset$class) <- make.names(levels(testset$class))

levels(preds_test) <- make.names(levels(preds_test)) 

confM_test <- table(Predetto = preds_test, Reale = testset$class)
print(confM_test)

conf_mat_df <- as.data.frame(confM_test)
colnames(conf_mat_df) <- c("Predicted", "Actual", "Freq")

class_levels <- levels(testset$class)
conf_mat_df$Actual<- factor(conf_mat_df$Actual, levels = rev(class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = class_levels)

ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") +
  coord_fixed() +
  labs(
    title = paste0("Confusion matrix SVM - CV (Test Set)\n",
                   "C = ", signif(best_C, 3), " | γ = ", signif(best_Gamma, 3)),
    x = "Predicted class",
    y = "Actual class"
  ) +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) + 
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"), 
    legend.position = "right" #
  )
```

### 3.2 Bo

```{r}
task <- makeClassifTask(data = dataset, target = "class")

learner <- makeLearner(
  "classif.svm",
  predict.type = "prob"
)

par.set <- makeParamSet(makeNumericParam("cost", lower = -2, upper = 2, trafo = function(x) 10 ^x),
                        makeDiscreteParam("kernel", values = c("linear", "radial")),
                        makeNumericParam("gamma", lower = -4, upper = 4, trafo = function(x) 10 ^x, requires = quote(kernel == "radial")))

ctrl <- makeMBOControl()
ctrl <- setMBOControlTermination(ctrl, iters = 30)
tune.ctrl <- makeTuneControlMBO(mbo.control = ctrl) 

run <- tuneParams(learner = learner, 
                  task = task, 
                  resampling = cv3, 
                  measures = acc, 
                  par.set = par.set, 
                  control = tune.ctrl)

ys <- getOptPathY(run$opt.path)

df <- data.frame(
  iter = seq_along(ys),
  value = ys,
  best_so_far = cummax(ys)
)

ggplot(df, aes(x = iter)) +
  geom_line(aes(y = value), color = "grey", linewidth = 1.2) +
  geom_point(aes(y = value), color = "black", size = 2) +
  geom_step(aes(y = best_so_far), color = "blue", linewidth = 1.4) +
  labs(
    title = "Optimization Path SVM - BO",
    x = "Iteration",
    y = "Objective Value"
  ) +
  theme_minimal(base_size = 14)

final_learner <- setHyperPars(learner, par.vals = run$x)
final_model <- mlr::train(final_learner, task)

preds <- predict(final_model, newdata = testset[,-ncol(testset)])$data$response
df_preds <- data.frame(Actual = testset$class, Predicted = preds)

levels(df_preds$Actual) <- make.names(levels(df_preds$Actual))
levels(df_preds$Predicted) <- make.names(levels(df_preds$Predicted))

conf_mat <- table(Predicted = df_preds$Predicted, Actual = df_preds$Actual)
conf_mat_df <- as.data.frame(conf_mat)

current_levels_actual <- levels(conf_mat_df$Actual)
current_levels_predicted <- levels(conf_mat_df$Predicted)

conf_mat_df$Actual    <- factor(conf_mat_df$Actual, levels = rev(c("X.1", "X1")))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = c("X.1", "X1"))


ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") + # Aggiunto name
  coord_fixed() +
  labs(
    title = paste0("Confusion matrix SVM - BO (Test Set)\n",
                   "C = ", round(run$x$cost, 3), " | γ = ", round(run$x$gamma, 3)),
    x = "Predicted class", 
    y = "Actual class"
  ) +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) + 
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

svm_model <- getLearnerModel(final_model)

# Numero assoluto di support vectors
n_sv <- svm_model$tot.nSV

# Numero totale di osservazioni nel training set usato da mlr
n_train <- getTaskSize(task)

# Percentuale
perc_sv <- n_sv / n_train * 100

n_sv
perc_sv

```

## 4. Gaussian Process

```{r}
set.seed(1)
dataset <- ds[ixs$Resample1, c(1,2, ncol(ds))]
testset <- ds[-ixs$Resample1, c(1,2, ncol(ds))]

X_train <- as.matrix(dataset[, -ncol(dataset)]) 
y_train_factor <- dataset$class
y_train <- as.numeric(as.character(y_train_factor))  # 1 e -1

test_label <- testset$class
testset <- testset[,-ncol(testset)]
X_test <- as.matrix(testset)
y_test <- as.numeric(as.character(test_label))

kern <- Gaussian$new(D = ncol(X_train))
gp_model <- GauPro_kernel_model$new(
  X = X_train,
  Z = y_train,
  kernel = kern,
  verbose = 1
)

# 5. Predizioni
test_preds <- gp_model$predict(X_test)
test_pred_class <- ifelse(test_preds > 0, 1, -1)

# Emp. Error (training)
train_preds <- gp_model$predict(X_train)
train_pred_class <- ifelse(train_preds > 0, 1, -1)

# 6. Accuracy
train_accuracy <- mean(train_pred_class == y_train)
test_accuracy <- mean(test_pred_class == y_test)

cat("Training Accuracy:", round(train_accuracy, 4), "\n")
cat("Test Accuracy:", round(test_accuracy, 4), "\n")

accuracy_df <- data.frame(
  Set = c("Training", "Test"),
  Accuracy = c(train_accuracy, test_accuracy)
)

conf_matrix <- t(table(Predicted = y_test, Actual = test_pred_class))
conf_df <- as.data.frame(conf_matrix)

conf_df$Actual <- factor(conf_df$Actual, levels = c(1, -1))
conf_df$Predicted <- factor(conf_df$Predicted, levels = c(-1, 1))

ggplot(conf_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Freq") +
  coord_fixed() +
  labs(
    title = "Confuson Matrix GP (Test Set)",
    x = "Predicted Class",
    y = "Actual class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

gp_model$kernel$beta
```

## 5. Random Forest

```{r}



data <- read.csv('/Users/francescofavagrossa/Desktop/PM.csv')

target <- read.csv('/Users/francescofavagrossa/Desktop/Buono_scarto.csv')

data_final <- cbind(data, target)


task <- makeClassifTask(
  id = "binary_classification",
  data = dataset,
  target = "class"
)

learner <- makeLearner(
  "classif.randomForest",
  predict.type = "prob"
)

param_set <- makeParamSet(
  makeIntegerParam("ntree", lower = 100, upper = 1000), # numero di alberi
  makeIntegerParam("mtry", lower = 1, upper = 12), # ho max 12 features
  makeIntegerParam("nodesize", lower = 1, upper = 20), # quanto lungo l'albero
  makeIntegerParam("maxnodes", lower = 10, upper = 100) # nodi collegati
)

rdesc <- makeResampleDesc("CV", iters = 10, stratify = TRUE)

mbo_ctrl <- makeMBOControl()
mbo_ctrl <- setMBOControlTermination(mbo_ctrl, iters = 30)
mbo_ctrl <- setMBOControlInfill(mbo_ctrl, crit = makeMBOInfillCritEI())

ctrl <- makeTuneControlMBO(
  mbo.control = mbo_ctrl,
  mbo.design = generateDesign(n = 10, par.set = param_set)
)

result <- tuneParams(
  learner = learner,
  task = task,
  resampling = rdesc,
  measures = list(acc),
  par.set = param_set,
  control = ctrl,
  show.info = TRUE
)

print(result)
cat("\nBest Hyperparameters:\n")
print(result$x)
cat("\nBest Performance:\n")
print(result$y)

final_learner <- setHyperPars(learner, par.vals = result$x) 

final_model <- mlr::train(final_learner, task) 

opt_path <- as.data.frame(result$opt.path) 


ggplot(opt_path, aes(x = seq_along(acc.test.mean), y = acc.test.mean)) +
  geom_line(color = "blue", alpha = 0.6) +
  geom_point(color = "darkblue", size = 3) +
  geom_hline(yintercept = max(opt_path$acc.test.mean), 
             linetype = "dashed", color = "red") +
  labs(title = "Random Forest Bayesian Optimization Progress",
       x = "Iteration", y = "Accuracy") +
  theme_minimal()

ggplot(opt_path, aes(x = ntree, y = mtry)) +
  geom_point(aes(color = acc.test.mean, size = acc.test.mean), alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Random Forest Parameter Space Exploration",
    x = "Number of Trees",
    y = "mtry",
    color = "Accuracy"
  ) +
  guides(size = "none") +
  theme_minimal()

ggplot(opt_path, aes(x = nodesize, y = acc.test.mean)) +
  geom_point(aes(color = mtry), size = 3) +
  geom_smooth(method = "loess", se = TRUE, color = "darkblue") +
  labs(title = "Effect of Node Size on Accuracy",
       x = "Node Size", y = "Accuracy") +
  theme_minimal()

#### FINAL MODEL PERFORMANCE ####

preds <- predict(final_model, newdata = testset)$data$response # <- TESTSET
df_preds <- data.frame(Actual = test_label, Predicted = preds)

conf_mat <- table(Predicted = df_preds$Predicted, Actual = df_preds$Actual)

conf_mat_df <- as.data.frame(conf_mat)

class_levels <- levels(conf_mat_df$Actual)
conf_mat_df$Actual    <- factor(conf_mat_df$Actual,    levels = rev(class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = class_levels)

 ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") +
  coord_fixed() +
  
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) + 
  labs(
    title = "Confusion matrix Random Forest - BO (Test Set)",
    x = "Predicted class",
    y = "Actual class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5), 
    legend.position = "right"
  )
```

## 6. Variable Selection

```{r}
set.seed(1)
dataset <- O_DS
X <- as.matrix(dataset[, -ncol(dataset)])
y <- dataset$class
y <- ifelse(y == -1, 0, 1) # -1 -> 0 ; 1 -> 1

cv <- cv.glmnet(X, y, family = "binomial", alpha = 1, nfolds = 10)
fit <- cv$glmnet.fit
beta_mat   <- as.matrix(fit$beta)
lambdas    <- fit$lambda 
feat_names <- rownames(beta_mat)

df <- as.data.frame(beta_mat)
colnames(df) <- paste0("lam_", seq_along(lambdas))
df$feature <- feat_names

df_long <- df |>
  pivot_longer(cols = starts_with("lam_"),
               names_to = "step", values_to = "coef") |>
  mutate(idx        = as.integer(sub("lam_", "", step)),
         lambda     = lambdas[idx],
         log_lambda = log(lambda)) |>
  select(-idx) |>
  group_by(feature) |>
  filter(any(coef != 0)) |>
  ungroup()

df_labels <- df_long %>%
  group_by(feature) %>%
  filter(lambda == min(lambda[coef != 0])) %>%
  ungroup()

ggplot(df_long, aes(x = log_lambda, y = coef, color = feature)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = log(cv$lambda.min), linetype = 2, color = "black") +
  geom_vline(xintercept = log(cv$lambda.1se), linetype = 2, color = "grey40") +
  labs(
    title = "LASSO Coefficient Paths (One Color per Variable)",
    x = "log(lambda)",
    y = "Coefficient",
    color = "Feature"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  ) +
  expand_limits(x = min(df_long$log_lambda) - 0.4)
```

\
\
\
\

### 6.1 KNN - BO with selected variable

```{r}
set.seed(1)
dataset <- ds[ixs$Resample1,c(1,2,ncol(ds)) ] # 90%
testset <- ds[-ixs$Resample1,c(1,2,ncol(ds)) ] # 10%
par.set <- makeParamSet(makeIntegerParam("k", lower = 1, upper = 51))

task <- makeClassifTask(data = dataset , target = "class")
ctrl <- makeMBOControl()
ctrl <- setMBOControlTermination(ctrl, iters = 30)
tune.ctrl <- makeTuneControlMBO(mbo.control = ctrl)

lrn <- makeLearner("classif.knn")

# tuning
run <- tuneParams(
  learner     = lrn,
  task        = task,
  resampling  = cv3,
  measures    = acc,
  par.set     = par.set,
  control     = tune.ctrl,
  show.info   = TRUE
)

ys <- getOptPathY(run$opt.path)
df <- data.frame(
  iter = seq_along(ys),
  value = ys,
  best_so_far = cummax(ys)
)
p1 <- ggplot(df, aes(x = iter)) +
  geom_line(aes(y = value), color = "gray", linewidth = 1.2) +
  geom_point(aes(y = value), color = "black", size = 2) +
  geom_step(aes(y = best_so_far), color = "blue", linewidth = 1.4) +
  labs(
    title = "Optimization Path KNN - BO",
    x = "Iteration",
    y = "Objective Value"
  ) +
  theme_minimal(base_size = 14) + 
  theme(plot.title = element_text(face = "bold"))

ks_df <- as.data.frame(getOptPathX(run$opt.path))
iters <- seq_len(nrow(ks_df))
accs <- getOptPathY(run$opt.path)

dfk <- data.frame(
  accuracy = ys,
  k = ks_df$k
)
dfk_sorted <- dfk %>% arrange(k)

miglior_k_bo <- run$x$k
final_learner <- setHyperPars(lrn, par.vals = run$x)
final_model <- mlr::train(final_learner, task)

# Predict on test set
preds <- predict(final_model, newdata = testset)$data$response

# Confusion matrix
conf_matrix <- confusionMatrix(preds, testset$class)
conf_mat_df <- as.data.frame(conf_matrix$table)
colnames(conf_mat_df) <- c("Predicted", "Actual", "Freq")

# Get class levels from testset
current_class_levels <- levels(testset$class)
conf_mat_df$Actual <- factor(conf_mat_df$Actual, levels = rev(current_class_levels))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = current_class_levels)

# Plot confusion matrix
p2 <- ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") +
  coord_fixed() +
  labs(
    title = paste0("Confusion Matrix KNN - BO (Lasso Selection)\n (Test Set) (k = ", miglior_k_bo, ") - Test Set"),
    x = "Predicted Class",
    y = "Actual Class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
grid.arrange(arrangeGrob(p1, bottom = "(a) Opt. Path "),
             arrangeGrob(p2, bottom = "(b) Confusion Matrix"), ncol = 2, nrow = 1)


```

### 6.2 SVM - BO with selected variables

```{r}
set.seed(2)
dataset <- ds[ixs$Resample1,c(1,2,ncol(ds)) ] # 90%
testset <- ds[-ixs$Resample1,c(1,2,ncol(ds)) ] # 10%
task <- makeClassifTask(data = dataset, target = "class")

learner <- makeLearner(
  "classif.svm",
  predict.type = "prob"
)

par.set <- makeParamSet(makeNumericParam("cost", lower = -2, upper = 2, trafo = function(x) 10 ^x),
                        makeDiscreteParam("kernel", values = c("linear", "radial")),
                        makeNumericParam("gamma", lower = -4, upper = 4, trafo = function(x) 10 ^x, requires = quote(kernel == "radial")))

ctrl <- makeMBOControl()
ctrl <- setMBOControlTermination(ctrl, iters = 30)
tune.ctrl <- makeTuneControlMBO(mbo.control = ctrl) 

run <- tuneParams(learner = learner, 
                  task = task, 
                  resampling = cv3, 
                  measures = acc, 
                  par.set = par.set, 
                  control = tune.ctrl)

ys <- getOptPathY(run$opt.path)

df <- data.frame(
  iter = seq_along(ys),
  value = ys,
  best_so_far = cummax(ys)
)

p1 <- ggplot(df, aes(x = iter)) +
  geom_line(aes(y = value), color = "grey", linewidth = 1.2) +
  geom_point(aes(y = value), color = "black", size = 2) +
  geom_step(aes(y = best_so_far), color = "blue", linewidth = 1.4) +
  labs(
    title = "Optimization Path SVM - BO (Lasso Selection)",
    x = "Iteration",
    y = "Objective Value"
  ) +
  theme_minimal(base_size = 14) + 
  theme(plot.title = element_text(face = "bold"))

final_learner <- setHyperPars(learner, par.vals = run$x)
final_model <- mlr::train(final_learner, task)

preds <- predict(final_model, newdata = testset[,-ncol(testset)])$data$response
df_preds <- data.frame(Actual = testset$class, Predicted = preds)

levels(df_preds$Actual) <- make.names(levels(df_preds$Actual))
levels(df_preds$Predicted) <- make.names(levels(df_preds$Predicted))

conf_mat <- table(Predicted = df_preds$Predicted, Actual = df_preds$Actual)
conf_mat_df <- as.data.frame(conf_mat)

current_levels_actual <- levels(conf_mat_df$Actual)
current_levels_predicted <- levels(conf_mat_df$Predicted)

conf_mat_df$Actual    <- factor(conf_mat_df$Actual, levels = rev(c("X.1", "X1")))
conf_mat_df$Predicted <- factor(conf_mat_df$Predicted, levels = c("X.1", "X1"))


p2 <- ggplot(conf_mat_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "skyblue2", high = "darkblue", name = "Freq") + # Aggiunto name
  coord_fixed() +
  labs(
    title = paste0("Confusion matrix SVM - BO (Lasso Selection)\n (Test Set)\n",
                   "C = ", round(run$x$cost, 3), " | γ = ", round(run$x$gamma, 3)),
    x = "Predicted class", 
    y = "Actual class"
  ) +
  scale_x_discrete(labels = c("X.1" = "-1", "X1" = "1")) +
  scale_y_discrete(labels = c("X1" = "1", "X.1" = "-1")) + 
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
grid.arrange(arrangeGrob(p1, bottom = "(a) Opt. Path "),
             arrangeGrob(p2, bottom = "(b) Confusion Matrix"), ncol = 2, nrow = 1)

svm_model <- getLearnerModel(final_model)
n_sv <- svm_model$tot.nSV
n_train <- getTaskSize(task)
perc_sv <- n_sv / n_train * 100
n_sv
perc_sv

```

## Plot Finale

```{r}
data_modelli <- data.frame(
  Modello = c(
    "KNN Hold Out", 
    "KNN Cross Validation", 
    "KNN Bayesian Optimization",
    "SVM Cross Validation",
    "SVM Bayesian Optimization", 
    "Gaussian Process", 
    "Random Forest Bayesian Optimization",
    "KNN Bayesian Optimization (Lasso) ",
    "SVM Bayesian Optimization (Lasso) "
  ),
  Accuracy = c(
    0.64, 0.70, 0.69, 0.685, 0.671, 1.00, 1.00, 1.00, 1.00
  )
)

# Converti in factor con l'ordine che vuoi
data_modelli$Modello <- factor(data_modelli$Modello, 
                                levels = c(
                                  "KNN Hold Out", 
                                  "KNN Cross Validation", 
                                  "KNN Bayesian Optimization",
                                  "SVM Cross Validation",
                                  "SVM Bayesian Optimization", 
                                  "Gaussian Process", 
                                  "Random Forest Bayesian Optimization",
                                  "KNN Bayesian Optimization (Lasso) ",
                                  "SVM Bayesian Optimization (Lasso) "
                                ))

ggplot(data_modelli, aes(x = Modello, y = Accuracy)) +
  geom_bar(stat = "identity", aes(fill = Accuracy)) + 
  labs(y = "Accuracy", 
       title = "Accuracy level compared between all the proposed models", 
       x = "Models") +
  scale_y_continuous(limits = c(0, 1.00), breaks = seq(0, 1.00, 0.25)) +
  scale_fill_gradient(low = "skyblue2", high = "darkblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(color = "gray80"),
        panel.grid.major.x = element_blank()
  )

```
